---
title: "Project 1 Brainstorm"
author: "Kristen Rose"
date: "2022-10-05"
output: html_document
---

# Loading Data and Packages

```{r}
diamonds <- read.csv("diamonds4.csv", header=TRUE)
library(tidyverse)
```

# Data Manipulation


```{r}
diamonds$cut.order <- factor(diamonds$cut)
diamonds$cut.order <- factor(diamonds$cut.order, c("Good", "Very Good", "Ideal", "Astor Ideal"))
new.levels<-c("Good or Very Good", "Good or Very Good", "Ideal or Astor Ideal", "Ideal or Astor Ideal")
cut2 <- factor(new.levels[diamonds$cut.order])
levels(cut2)
diamonds <- data.frame(diamonds, cut2)
```

```{r}
diamonds$color.order <- factor(diamonds$color)
diamonds$color.order <- factor(diamonds$color.order, c("J", "I", "H", "G", "F", "E", "D"))
new.levels2<-c('Near-colorless','Near-colorless','Near-colorless','Near-colorless','Colorless','Colorless','Colorless')
color2 <- factor(new.levels2[diamonds$color.order])
levels(color2)
diamonds <- data.frame(diamonds, color2)
```

```{r}
diamonds$clarity.order <- factor(diamonds$clarity)
diamonds$clarity.order <- factor(diamonds$clarity.order, c("SI1","SI2","VS1","VS2","VVS1","VVS2","IF","FL"))
new.levels3 <- c("Slightly Included", "Slightly Included", "Very Slightly Included", "Very Slightly Included", "Very, Very Slightly Included", "Very, Very Slightly Included", "Internally Flawless or Flawless", "Internally Flawless or Flawless")
clarity2 <- factor(new.levels3[diamonds$clarity.order])
levels(clarity2)
diamonds <- data.frame(diamonds, clarity2)
diamonds$clarity2 <- factor(diamonds$clarity2, c("Slightly Included", "Very Slightly Included", "Very, Very Slightly Included", "Internally Flawless or Flawless"))
```

```{r}
carat2 <- ifelse(diamonds$carat<2.0, "< 2.0 carats", ">= 2.0 carats")
diamonds <- data.frame(diamonds, carat2)
```
```{r}
median(diamonds$carat)
```

```{r}
carat3<-cut(diamonds$carat, breaks = c(-Inf, 1, 2, Inf),
labels = c("Less than 1 carat", "1-2 carats", "More than 2 carats"))
```

# Data Visualizations

**How is price related to other variables?**

Price and Carat

```{r}
diamonds$log.price <- log(diamonds$price)

ggplot(diamonds, aes(x=carat,y=log.price))+
  geom_point()
```

Price and Clarity

```{r}
ggplot(diamonds, aes(x=clarity2, y=log.price))+
  geom_boxplot(color="black", outlier.color = "gray")
```

Price and Color

```{r}
ggplot(diamonds, aes(x=color2, y=log.price))+
  geom_boxplot(color="black", outlier.color = "yellow")
```

Price and Cut

```{r}
ggplot(diamonds, aes(x=cut2, y=log.price))+
  geom_boxplot(color="black", outlier.color = "green")
```

**How do other variables relate to each other?**


```{r}
ggplot(diamonds,aes(x=clarity2, fill=cut2))+
geom_bar(position = "fill")+
labs(x="Clarity", y="Count", title="Cut of Diamond by Clarity", fill="Cut")

```

- As the clarity improves, the diamond is more likely to have an Ideal or Astor Ideal cut.

```{r}
ggplot(diamonds, aes(x=carat2, y=log.price, fill=cut2))+
geom_boxplot()+
labs(x="Carat", y="log.price", fill="Cut",
title="Distribution of Price by Cut & Carat")
```

- Interesting to note in the above graph that for smaller diamonds, ideal cuts have a slightly lower median price than good cuts.

```{r}
ggplot(diamonds,aes(x=carat2, fill=cut2))+
geom_bar()+
facet_wrap(~clarity2)+
labs(x="Carat", y="Count", fill="Cut")
```

- Interesting to note in the above graph that there tend to be a greater percentage of ideal cuts for smaller rings than larger diamonds - for both colorless and near-colorless diamonds. 

# Simple Linear Regression

```{r}
ggplot(diamonds, aes(x=carat,y=price))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="Carat", y="Price", title="Scatterplot of Price Against Carat")
```

```{r}
result<-lm(price~carat, data=diamonds)

##store fitted y & residuals
yhat<-result$fitted.values
res<-result$residuals

##add to data frame
diamonds<-data.frame(diamonds,yhat,res)

##residual plot
ggplot(diamonds, aes(x=yhat,y=res))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="Residual Plot")
```

```{r}
library(MASS)
boxcox(result, lambda = seq(0, 0.5, 1/10))
```

```{r}
ystar<-log(diamonds$price)
diamonds<-data.frame(diamonds,ystar)
result.ystar<-lm(ystar~carat, data=diamonds)

##store fitted y & residuals
yhat2<-result.ystar$fitted.values
res2<-result.ystar$residuals

##add to data frame
diamonds<-data.frame(diamonds,yhat2,res2)

##residual plot with ystar
ggplot(diamonds, aes(x=yhat2,y=res2))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="Residual Plot with ystar")
```


```{r}
ggplot(diamonds, aes(x=carat,y=ystar))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="Carat", y="Price", title="Scatterplot of Transformed Price Against Carat")
```

```{r}
xstar = log(diamonds$carat)
diamonds = data.frame(diamonds,xstar)
result.xstar<-lm(ystar~xstar, data=diamonds)

yhat3<-result.xstar$fitted.values
res3<-result.xstar$residuals

##add to data frame
diamonds<-data.frame(diamonds,yhat3,res3)

##residual plot with xstar
ggplot(diamonds, aes(x=yhat3,y=res3))+
  geom_point()+
  geom_hline(yintercept=0, color="red")+
  labs(x="Fitted y", y="Residuals", title="Residual Plot with xstar")
```
```{r}
ggplot(diamonds, aes(x=xstar,y=ystar))+
  geom_point()+
  geom_smooth(method = "lm", se=FALSE)+
  labs(x="Carat", y="Price", title="Scatterplot of Transformed Price Against Carat")
```

